---
title: "Random Number Generation"
subtitle: "HW 6 of STAT 5361 Statistical Computing"
author: Biju Wang^[<bijuwang@uconn.edu>]
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
header-includes: 
  \usepackage{float}
  \usepackage{algorithm}
  \usepackage{algpseudocode}
  \floatplacement{figure}{H}
output: 
  pdf_document:
    number_sections: true
---

# Rejecion Sampling
## Normalizing Constant for $g$
Since we have the following intergral
\begin{align*}
\int^{\infty}_{0}(2x^{\theta-1}+x^{\theta-\frac{1}{2}})e^{-x}dx & =2\int^{\infty}_{0}x^{\theta-1}e^{-x}dx+\int^{\infty}_{0}x^{\theta-\frac{1}{2}}e^{-x}dx\\
   & =2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})
\end{align*}
Obviously
$$C=\frac{1}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}$$
Therefore
$$g(x)=\frac{2\Gamma(\theta)}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}\frac{1}{\Gamma(\theta)}x^{\theta-1}e^{-x}+\frac{\Gamma(\theta+\frac{1}{2})}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}\frac{1}{\Gamma(\theta+\frac{1}{2})}x^{\theta-\frac{1}{2}}e^{-x}$$
Thus, $g(x)$ is a mixuture of Gamma distributions. The component distributions are Gamma$(\theta, 1)$ and Gamma$(\theta+\frac{1}{2}, 1)$, the corresponding weights are $\frac{2\Gamma(\theta)}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}$ and $\frac{\Gamma(\theta+\frac{1}{2})}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}$.

## Sampling from $g(x)$
The pseudo-code is as follows
\begin{algorithm}
\begin{algorithmic}[1]
\Procedure{My Procedure}{}
\State Sample $U$ from U$(0, 1)$
\If{$U < \frac{2\Gamma(\theta)}{2\Gamma(\theta)+\Gamma(\theta+\frac{1}{2})}$}
\State Sample $X$ from Gamma$(\theta, 1)$
\Else
\State Sample $X$ from Gamma$(\theta+\frac{1}{2}, 1)$
\EndIf
\State\Return $X$
\EndProcedure
\end{algorithmic}
\caption{Sampling from $g(x)$}
\end{algorithm}

```{r, message = FALSE, fig.asp=.5, fig.pos = "H", fig.cap = "Histogram, Kernel Curve and True Curve for Generated Samples from $g(x)$ with $\\theta=1$"}
sample.g <- function(n, shape, scale, prob){
  x <- rep(0, n)
  u <- runif(n, 0, 1)
  g1 <- rgamma(n, shape = shape[1], scale = scale[1])
  g2 <- rgamma(n, shape = shape[2], scale = scale[2])
  
  x[u < prob] <- g1[u < prob]
  x[u >= prob] <- g2[u >= prob]
  
  x
}

theta <- 1
prob <- 2 * gamma(theta)/(2 * gamma(theta) + gamma(theta + 1/2))
shape <- c(theta, theta + 1/2)
scale <- c(1, 1)
n <- 10000

g <- function(x, theta){
    (2 * x^(theta - 1) * exp(-x) + x^(theta - 1/2) * exp(-x))/(2 * gamma(theta) + gamma(theta + 1/2))
}

x <- sample.g(n, shape, scale, prob)

library("ggplot2")
ggplot(data.frame(x = x), aes(x = x)) +
geom_histogram(aes(y=..density..), color = "black", alpha = 0.2) +
geom_density(aes(x, color = "a")) +
stat_function(aes(x, color = "b"), fun = function(x) g(x, theta = theta)) +
scale_colour_manual(name = "Legend", values = c("a" = "blue", "b" = "red"),
                    labels = c("Kernel", "True")) +
labs(x = expression("Values of"~x), y = expression("Density of"~g(x))) +
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle(expression("Gamma Mixture Distribution with"~theta==1))
```

## Sampling from $f(x)$ via Rejection Sampling
Since our target density is $f(x)$ and $f(x)\varpropto q(x)=\sqrt{4+x}x^{\theta-1}e^{-x}$ and the instrumental density is $g(x)$, we first calculate the minimal $\alpha$ such that $q(x)\leqslant\alpha g(x)$.
\begin{align*}
\alpha & =\sup_{x>0}\frac{q(x)}{g(x)}\\
       & =\sup_{x>0}\frac{\sqrt{4+x}x^{\theta-1}e^{-x}}{C(2x^{\theta-1}+x^{\theta-\frac{1}{2}})e^{-x}}\\
       & =\sup_{x>0}\frac{\sqrt{4+x}}{C\sqrt{x}}\\
       & =\frac{1}{C}
\end{align*}
Therefore
$$q(x)=\sqrt{4+x}x^{\theta-1}e^{-x}\leqslant \alpha g(x)=\frac{1}{C}g(x)=(2x^{\theta-1}+x^{\theta-\frac{1}{2}})e^{-x}$$
The pseudo-code is as follows
\begin{algorithm}
\begin{algorithmic}[1]
\Procedure{My Procedure}{}
\State Sample $X$ from $g(x)$ and $U$ from U$(0, 1)$\label{sample}
\If{$U \leqslant \frac{q(X)}{\alpha g(X)}$}
\State\Return $X$
\Else
\State Go back to \ref{sample}
\EndIf
\EndProcedure
\end{algorithmic}
\caption{Sampling from $f(x)$}
\end{algorithm}

```{r, message = FALSE, fig.asp=.5, fig.pos = "H", fig.cap = "Histogram, Kernel Curve and True Curve for Generated Samples from $f(x)$ with $\\theta=1$"}
sample.f <- function(n, shape, scale, prob){
  sample <- rep(0, n)
  iter <- 1
  
  while (iter <= n) {
      x <- sample.g(1, shape, scale, prob)
      u <- runif(1, 0, 1)
      if(u <= sqrt(4 + x)/(2 + sqrt(x))){
        sample[iter] <- x
        iter <- iter + 1
      } 
  }
  sample
}

theta <- 1
prob <- 2 * gamma(theta)/(2 * gamma(theta) + gamma(theta + 1/2))
shape <- c(theta, theta + 1/2)
scale <- c(1, 1)
n <- 10000

h <- function(x, theta){
  sqrt(4 + x) * x^(theta - 1) * exp(-x)
}
integral <- integrate(function(x) h(x, theta = theta), 0, Inf)

f <- function(x, theta){
  sqrt(4 + x) * x^(theta - 1) * exp(-x)/integral$value
}

x <- sample.f(n, shape, scale, prob)

library("ggplot2")
ggplot(data.frame(x = x), aes(x = x)) +
geom_histogram(aes(y=..density..), color = "black", alpha = 0.2) +
geom_density(aes(x, color = "a")) +
stat_function(aes(x, color = "b"), fun = function(x) f(x, theta = theta)) +
scale_colour_manual(name = "Legend", values = c("a" = "blue", "b" = "red"),
                    labels = c("Kernel", "True")) +
labs(x = expression("Values of"~x), y = expression("Density of"~f(x))) +
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle(expression("Distribution of"~f(x)~"with"~theta==1))
```


# Mixture of Beta
## Treating $f(x)$ as a Whole and Using Rejection Sampling
It's easy to see
$$q(x)=\frac{x^{\theta-1}}{1+x^{2}}+\sqrt{2+x^{2}}(1-x)^{\beta-1}\leqslant x^{\theta-1}+2(1-x)^{\beta-1}$$
One intuitive way to choose mixed Beta distributions(here we choose $\lambda=\frac{1}{2}$) is
$$g(x)=\lambda \text{Beta}(\theta, 1)+(1-\lambda)\text{Beta}(1, \beta)=\frac{\lambda}{\theta}x^{\theta-1}+\frac{1-\lambda}{\beta}(1-x)^{\beta-1}=\frac{1}{2\theta}x^{\theta-1}+\frac{1}{2\beta}(1-x)^{\beta-1}$$
And select $\alpha=\max\{2\theta, 4\beta\}$. The following inequality is straightforward
$$q(x)=\frac{x^{\theta-1}}{1+x^{2}}+\sqrt{2+x^{2}}(1-x)^{\beta-1}\leqslant x^{\theta-1}+2(1-x)^{\beta-1}\leqslant\alpha g(x)$$
The pseudo-code is as follows
\begin{algorithm}
\begin{algorithmic}[1]
\Procedure{My Procedure}{}
\State Sample $U$ from U$(0, 1)$\label{sample}
\If{$U<\lambda$}
\State Sample $X$ from Beta$(\theta, 1)$
\Else
\State Sample $X$ from Beta$(1, \beta)$
\EndIf

\State Sample $U_{1}$ from U$(0, 1)$
\If{$U_{1}\leqslant\frac{q(X)}{\alpha g(X)}$}
\State\Return $X$
\Else
\State Go back to \ref{sample}
\EndIf
\EndProcedure
\end{algorithmic}
\caption{Sampling from $f(x)$ with treating $f(x)$ as a whole}
\end{algorithm}

```{r, message = FALSE, fig.asp=.4, fig.pos = "H", fig.cap = "Histogram, Kernel Curve and True Curve for Generated Samples from $f(x)$ with $\\theta=2$, $\\beta=2$"}
sample.g <- function(n, shape1, shape2, prob){
  x <- rep(0, n)
  u <- runif(n, 0, 1)
  g1 <- rbeta(n, shape1 = shape1[1], shape2 = shape2[1])
  g2 <- rbeta(n, shape1 = shape1[2], shape2 = shape2[2])
  
  x[u < prob] <- g1[u < prob]
  x[u >= prob] <- g2[u >= prob]
  
  x
}

sample.f <- function(n, shape1, shape2, prob){
  sample <- rep(0, n)
  iter <- 1
  alpha <- max(shape1[1] * 2, shape2[2] * 4)
  
  while (iter <= n) {
      x <- sample.g(1, shape1, shape2, prob)
      u <- runif(1, 0, 1)
      
      q <- x ^ (shape1[1] - 1)/(1 + x^2) + sqrt(2 + x^2) * (1 - x)^(shape2[2] - 1)
      g <- (1/2) * dbeta(x, shape1 = shape1[1], shape2 = shape2[1]) + 
        (1/2) * dbeta(x, shape1 = shape1[2], shape2 = shape2[2])
      
      if(u <= q/(alpha * g)){
        sample[iter] <- x
        iter <- iter + 1
      } 
  }
  sample
}

theta <- 2
beta <- 2
prob <- 1/2

shape1 <- c(theta, 1)
shape2 <- c(1, beta)
n <- 10000


h <- function(x, theta, beta){
  x ^ (theta - 1)/(1 + x^2) + sqrt(2 + x^2) * (1 - x)^(beta - 1)
}
integral <- integrate(function(x) h(x, theta = theta, beta = beta), 0, 1)

f <- function(x, theta, beta){
  (x ^ (theta - 1)/(1 + x^2) + sqrt(2 + x^2) * (1 - x)^(beta - 1))/integral$value
}

x <- sample.f(n, shape1, shape2, prob)

library("ggplot2")
ggplot(data.frame(x = x), aes(x = x)) +
geom_histogram(aes(y=..density..), color = "black", alpha = 0.2) +
geom_density(aes(x, color = "a")) +
stat_function(aes(x, color = "b"), fun = function(x) f(x, theta = theta, beta = beta)) +
scale_colour_manual(name = "Legend", values = c("a" = "blue", "b" = "red"),
                    labels = c("Kernel", "True")) +
labs(x = expression("Values of"~x), y = expression("Density of"~f(x))) +
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle(expression(paste(" Distribution of ", f(x), " with ", theta==2, ", ", beta==2 )))
```

## Treating $f(x)$ as the Sum of Two Functions and Using Rejection Sampling
Let's define
\begin{align*}
q_{1}(x) & =\frac{x^{\theta-1}}{1+x^{2}} & q_{2}(x) & =\sqrt{2+x^{2}}(1-x)^{\beta-1}\\
g_{1}(x) & =\frac{1}{\theta}x^{\theta-1} & g_{2}(x) & =\frac{1}{\beta}(1-x)^{\beta-1}\\
\alpha_{1} & =\theta & \alpha_{2} & =\sqrt{3}\beta
\end{align*}

Thus
\begin{gather*}
f(x)\varpropto q_{1}(x)+q_{2}(x)\\
q_{1}(x)\leqslant \alpha_{1}g_{1}(x) \qquad q_{2}\leqslant \alpha_{2}g_{2}(x)\\
\end{gather*}

The pseudo-code is as follows
\begin{algorithm}
\begin{algorithmic}[1]
\Procedure{My Procedure}{}
\State Sample $U$ from U$(0, 1)$\label{sample}

\If{$U<\frac{\alpha_{1}}{\alpha_{1}+\alpha_{2}}$}
\State Sample $X$ from $g_{1}(x)$ and $U_{1}$ from U$(0, 1)$
\If{$U_{1}\leqslant\frac{q_{1}(X)}{\alpha_{1}g_{1}(X)}$}
\State\Return $X$
\Else
\State Go back to \ref{sample}
\EndIf

\Else
\State Sample $X$ from $g_{2}(x)$ and $U_{2}$ from U$(0, 1)$
\If{$U_{2}\leqslant\frac{q_{2}(X)}{\alpha_{2}g_{2}(X)}$}
\State\Return $X$
\Else
\State Go back to \ref{sample}
\EndIf

\EndIf
\EndProcedure
\end{algorithmic}
\caption{Sampling from $f(x)$ with treating $f(x)$ as the sum of two functions}
\end{algorithm}

```{r, message = FALSE, fig.asp=.4, fig.pos = "H", fig.cap = "Histogram, Kernel Curve and True Curve for Generated Samples from $f(x)$ with $\\theta=2$, $\\beta=2$"}
sample.f <- function(n, shape1, shape2){
  sample <- rep(0, n)
  iter <- 1
  alpha1 <- shape1[1]
  alpha2 <- sqrt(3) * shape2[2]
  prob <- alpha1/(alpha1 + alpha2)
  
  while (iter <= n) {
    u <- runif(1, 0, 1)
    
    if(u < prob){
      x <- rbeta(1, shape1 = shape1[1], shape2 = shape2[1])
      u1 <- runif(1, 0, 1)
      
      if(u1 <= 1/(1 + x^2)){
        sample[iter] <- x
        iter <- iter + 1
      }
    }
    else{
      x <- rbeta(1, shape1 = shape1[2], shape2 = shape2[2])
      u2 <- runif(1, 0, 1)
      
      if(u2 <= sqrt(2 + x^2)/sqrt(3)){
        sample[iter] <- x
        iter <- iter + 1
      }
    }
  }
  sample
}

theta <- 2
beta <- 2

shape1 <- c(theta, 1)
shape2 <- c(1, beta)
n <- 10000


h <- function(x, theta, beta){
  x ^ (theta - 1)/(1 + x^2) + sqrt(2 + x^2) * (1 - x)^(beta - 1)
}
integral <- integrate(function(x) h(x, theta = theta, beta = beta), 0, 1)

f <- function(x, theta, beta){
  (x ^ (theta - 1)/(1 + x^2) + sqrt(2 + x^2) * (1 - x)^(beta - 1))/integral$value
}

x <- sample.f(n, shape1, shape2)

library("ggplot2")
ggplot(data.frame(x = x), aes(x = x)) +
geom_histogram(aes(y=..density..), color = "black", alpha = 0.2) +
geom_density(aes(x, color = "a")) +
stat_function(aes(x, color = "b"), fun = function(x) f(x, theta = theta, beta = beta)) +
scale_colour_manual(name = "Legend", values = c("a" = "blue", "b" = "red"),
                    labels = c("Kernel", "True")) +
labs(x = expression("Values of"~x), y = expression("Density of"~f(x))) +
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle(expression(paste(" Distribution of ", f(x), " with ", theta==2, ", ", beta==2 )))
```




